---
title: "Introduction to BIRDS"
author: "Alejandro Ruete and Debora Arlt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: TRUE
vignette: >
  %\VignetteIndexEntry{Introduction to BIRDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Installing `BIRDS`

The package `BIRDS` is currently available on GitHub and you can freely download 
the source code here [BIRDS on GitHub](https://github.com/Greensway/BIRDS)

The easiest option is to install the package directly from GitHub using the 
package `devtools`.  Install `devtools` if you do not already have installed it.
`devtools` may have some other dependencies. Please visits the following webpages 
to be sure you are on the right track:

* [devtools](https://www.r-project.org/nosvn/pandoc/devtools.html)  
* [Rtools](https://cran.r-project.org/bin/windows/Rtools/) 

```{r install package, eval = F}
install.packages('devtools')
library(devtools)
install_github('Greensway/BIRDS')
```

If you receive an error we would love to receive your bug report 
[here](https://github.com/Greensway/BIRDS/issues)

## Basic example and data requirements

Starting with a primary biodiversity data set (species observations) `BIRDS` employs 
three basic steps: organizing the data into a sampling event-based format, summarizing
the data producing summary variables that inform about sampling effort and 
data completeness, and reviewing the summarized variables.

This package works with `data.frame` tables containing raw species observations 
as rows (i.e. primary biodiversity data, PBD), with minimal information required
for each observation in the following columns:

1. species identification 
2. x and y spatial coordinates in two columns (epsg:4326 assumed if not otherwise 
specified)
3. date of the observation (either in one formatted column or three columns y-m-d)

The function `organiseBirds()` converts ‘data.frame’ into a `SpatialPointsDataFrame`
adding to each observation a unique identifier for the assumed visit (sampling event) 
it belongs to, given some input parameters. The function `organizeBirds()` will interpret 
the DarwinCore standard for column names as default, but other names can be specified 
too. For more details on how a visit is defined and options for each function please
refer to the [technical vignette](https://greensway.github.io/BIRDS/articles/technical_details.html) and to 
each function's help page.  

Then, the function `summariseBirds()` will overlay the data with a given spatial
grid and create a set of objects that summarize the data spatially, temporally, 
and spatio-temporally, and also provide other intermediate results useful for 
later analyses. Finally, the function `exportBirds()` helps the user to obtain 
the data ready to be plotted with widely known functions. 

We use as an example a data set consisting of 10k species observations of 
bumblebees (i.e. *Bombus* spp.) in Götaland (the southern part of Sweden) during
2000-2018. The data set `bombusObs` is part of this package and its metadata
can be found in the data set help page `?bombusObs`.  

You can easily create a grid over a sample area (i.e. `gotaland`), organize and 
summarize the data, and export the variables you want to plot:

```{r basic example, eval = TRUE}
library(BIRDS)
# Create a grid for your sample area that will be used to summarise the data:
grid <- makeGrid(gotaland, gridSize = 10)
# The grid can be easily created in different ways. 

# Import the species observation data:
PBD<-bombusObs
# alternatively, you could load a previously downloaded .CSV file 
# PBD <- read.csv(file="path/to/your/file.csv)

# Convert the data from an observation-based to a visit-based format, adding a unique identifier for each visit:
OB <- organizeBirds(PBD, sppCol = "scientificName", simplifySppName = TRUE)

# Summarise the data:
SB <- summariseBirds(OB, grid=grid)

# Look at some summarised variables:
# Number of observations
EBnObs <- exportBirds(SB, dimension = "temporal", timeRes = "yearly", 
                      variable = "nObs", method = "sum")
# Number of visits
EBnVis <- exportBirds(SB, dimension = "temporal", timeRes = "yearly", 
                      variable = "nVis", method = "sum")
# Average species list length (SLL) per year (a double-average, i.e. the mean over cell values
# for the median SLL from all visits per year and cell) 
EBavgSll <- colMeans(SB$spatioTemporal[,,"Yearly","avgSll"], na.rm = TRUE)
# The ratio of number of observations over number of visits
relObs<-EBnObs/EBnVis
```

Then, with functions you already know, plot this into a time series.
```{r figure 1, fig.show='hold', fig.width= 7, fig.height= 5, fig.cap = "Time series for *Bombus* spp. dataset."}
par(mar=c(4,4,1,6), las=1)
plot(time(EBnObs), EBnObs, type = "l", lwd = 3, xlab = "Year", ylab = "Number", 
     ylim=c(0, max(EBnObs)), xaxp=c(2000, 2018, 18))
lines(time(EBnObs), EBnVis, lwd=3, lty=2)
lines(time(EBnObs), relObs * max(EBnObs) / max(relObs), lwd=3, lty=1, col="#78D2EB")
lines(time(EBnObs), EBavgSll * max(EBnObs) / max(EBavgSll), lwd=3, lty=1, col="#FFB3B5")
axis(4, at = seq(0, max(EBnObs), length.out = 5), 
     labels = round(seq(0,max(relObs), length.out = 5), 1), 
     lwd = 2, col = "#78D2EB", col.ticks = "#78D2EB")
axis(4, at = seq(0, max(EBnObs), length.out = 5) , 
     labels = round(seq(0,max(EBavgSll), length.out = 5), 1), 
     lwd = 2, col = "#FFB3B5", col.ticks = "#FFB3B5", line = 3)
legend("topleft", legend=c("n.observations","n.visits"), 
       lty = c(1,2), lwd = 3, bty = "n")
legend("bottomright", legend=c("n.observations / n.visits", "avg. SLL per cell"),
       lty = 1, lwd = 3, col = c("#78D2EB", "#FFB3B5"), bty = "n")
```  

Although we downloaded the data in April 2019, we know that the latest reports
for 2018 were not yet uploaded and we see that in the figure.  
We can see that over the years there are on average more observations, more visits and 
also more species reported for each visit. While it looks like as sampling effort has increased
we can also see that there are on average fewer observations per visit. This suggests that the 
sampling effort per visit has actually decreased. There seem to be many more visits reported
in later years driving also an increase in number of observations, but many visits hold only few observations.
There seem to be also few visits with long species lists driving the increase of average SLL.

Then, we can summarize spatial data in some maps, to review, for example, 
Where do we have visits at all?, or How often are visits performed in July vs December?  

```{r figure 2, fig.show='hold', fig.width= 7, fig.height= 3}
library(sp)
wNonEmpty<-unname( which( unlist(lapply(SB$overlaid, nrow)) != 0) )
EB <- exportBirds(SB, "Spatial", "Month", "nYears", "sum")
# because the dimension is "spatial", the result is a 'SpatialPolygonDataFrame'

palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, max(EB@data, na.rm = TRUE)), 
                               na.color = "transparent")
par(mfrow=c(1,3), mar=c(1,1,1,1))
plot(SB$spatial[wNonEmpty,], col="grey", border = NA)
plot(gotaland, col=NA, border = "grey", lwd=1, add=TRUE)
mtext("Visited cells", 3, line=-1)

plot(EB, col=palBW(EB@data$Jul), border = NA)
plot(gotaland, col=NA, border = "grey", lwd=1, add=TRUE)
mtext("Number of years for which \nJuly was sampled", 3, line=-2)

plot(EB, col=palBW(EB@data$Dec), border = NA)
plot(gotaland, col=NA, border = "grey", lwd=1, add=TRUE)
mtext("Number of years for which \nDecember was sampled", 3, line=-2)
legend("bottomleft", legend=seq(0, max(EB@data, na.rm = TRUE), length.out = 5),
      col = palBW(seq(0, max(EB@data, na.rm = TRUE), length.out = 5)),
      title = "Number of years", pch = 15, bty="n")
```
  

We could also map the ignorance scores based on the either the number of observations 
or visits using the function `exposeIgnorance()`. Ignorance scores are a proxy 
for the lack of sampling effort, computed by making the number of observations 
relative to a reference number of observations that is considered to be enough to 
reduce the ignorance score by half (henceforth the Half-ignorance approach).
The algorithm behind the Ignorance Score is designed for comparison of bias and 
gaps in primary biodiversity data across taxonomy, time and space
 Read more here: Ruete 2015 Biodiv Data J 3:e5361, doi:10.3897/BDJ.3.e5361
 
```{r figure 3, fig.show='hold', fig.width= 7, fig.height= 4}
par(mfrow=c(1,2), mar=c(1,1,1,1))
palBW <- leaflet::colorNumeric(c("white", "navyblue"), 
                               c(0, max(SB$spatial@data$nVis, na.rm = TRUE)), 
                               na.color = "transparent")
seqNVis<-round(seq(0, max(SB$spatial@data$nVis, na.rm = TRUE), length.out = 5))
plot(SB$spatial, col=palBW(SB$spatial@data$nVis), border = NA)
plot(gotaland, col=NA, border = "grey", lwd=1, add=TRUE)
legend("bottomleft", legend=seqNVis, col = palBW(seqNVis),
      title = "Number of \nobservations", pch = 15, bty="n")

ign<-exposeIgnorance(SB$spatial@data$nVis, h = 5)
palBWR <- leaflet::colorNumeric(c("navyblue", "white","red"), c(0, 1), 
                                na.color = "transparent")
plot(gotaland, col="grey90", border = "grey90", lwd=1)
plot(SB$spatial, col=palBWR(ign), border = NA, add=TRUE)
plot(gotaland, col=NA, border = "grey", lwd=1, add=TRUE)
legend("bottomleft", legend=c(seq(0, 1, length.out = 5), "NA"),
      col = c(palBWR(seq(0, 1, length.out = 5)), "grey90"),
      title = "Ignorance \nnVis, \nO0.5=5", pch = 15, bty="n")
```  
  
The ignorance scores are highlighting where we have little information and hence cannot be 
confident about the lack of reports implying an absence of the species. 

<!-- Or we could connect to other packages like `vegan` to continue with community analyses. -->
<!-- In this case, we create a community matrix using the visits as sampling units,  -->
<!-- summarizing for each grid cell the number of visits a species was observed (i.e.  -->
<!-- ignoring replicates within visits). -->

<!-- ```{r community richness, fig.show='hold', fig.width=7, message=FALSE} -->
<!-- ## Community analysis -->
<!-- CM <- communityMatrix(SB, sampleUnit="visit") -->
<!-- library(vegan) -->
<!-- sp1 <- specaccum(CM, method = "exact") -->
<!-- par(mar=c(4,4,1,1), xlog=TRUE, las=1) -->
<!-- plot(sp1, ci.type="poly", col="blue", lwd=2, ci.lty=0,  -->
<!--      ci.col="lightblue", log="x",   -->
<!--      xlab="Number of grid cells sampled",  -->
<!--      ylab="Species richness",  -->
<!--      main="Species accumulation curve") -->
<!-- ``` -->
